Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-03-01 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: discussion on cs results

\newcommand{\h}{\mathpzc{h}}
\newcommand{\Hhat}{\hat{\mathpzc{H}}}
\newcommand{\B}{\mathpzc{B}}
\newcommand{\hlm}{\mathpzc{h}_{\ell m}}
\newcommand{\xilm}{\xi_{\ell m}}
\newcommand{\Ylm}{{Y}^{-2}_{\ell m}}
\newcommand{\Y}{{Y}^{-2}}
\newcommand{\hc}{h_\times}
\newcommand{\hp}{h_+}
\newcommand{\Fc}{F_\times}
\newcommand{\Fp}{F_+}
\newcommand{\Mf}{M_f}
\newcommand{\cA}{\mathpzc{A}}
\newcommand{\lm}{_{\ell m}}
\newcommand{\deff}{d_\mathrm{eff}}
\newcommand{\rmi}{\mathrm{i}}
\newcommand{\blambda}{\bm{\lambda}}
\newcommand{\btheta}{\bm{\theta}}
\newcommand{\bxi}{\bm{\xi}}
\newcommand{\bxigr}{\bm{\xi}_{\text{GR}}}
\newcommand{\bxingr}{\bm{\xi}_{\text{nGR}}}
\newcommand{\bzeta}{\bm{\zeta}}
\newcommand{\bs}[1]{\bm{\vec{S}_{#1}}}
\newcommand{\Mo}{M_{\odot}}
\newcommand{\FFe}{\mathrm{FF}_\mathrm{eff}}
\newcommand{\FF}{\mathrm{FF}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\rhoopt}{\rho_\mathrm{opt}}
\newcommand{\rhosubopt}{\rho_\mathrm{subopt}}
\newcommand{\fqnm}{f}
\newcommand{\sigmaqnm}{\sigma}
\newcommand{\n}{\mathbf{n}}
\newcommand*{\skymapscale}{0.5}
\newcommand*{\paramestscale}{0.455}
\newcommand{\df}[1]{\delta f_{\text{#1}}}
\newcommand{\dtau}[1]{\delta \tau_{\text{#1}}}
\newcommand{\fngr}[1]{f_{\text{#1}}}
\newcommand{\taungr}[1]{\tau_{\text{#1}}}
\newcommand{\fgr}[1]{f ^{\text{GR}}_{\text{#1}}}
\newcommand{\taugr}[1]{\tau ^{\text{GR}}_{\text{#1}}}
\newcommand{\pSEOB}{\texttt{pSEOBNR}}
\newcommand{\SEOB}{\texttt{SEOBNR}}

@@

There are two requirements that we must take into account before
we can confidently claim to have placed a constraint, with our model assumptions,
with our parameter estimation study.

First, as described in Sec.~\ref{sec:review_theories}, all theories that we consider must be considered as an
effective field theory, meaning it has a sensible description only below an energy, or equivalently an length scale.
%
As a threshold for the validity of the EFT we use
%
\begin{equation}
T_{\rm EFT} (\varepsilon; \, \mathfrak{m}) = \varepsilon \,  (G \mathfrak{m} / c^2)\,,
\end{equation}
%
where $\varepsilon$ is a dimensionless number and $\mathfrak{m}$ the median value of one of the
masses involved in the problem.
%
We also momentarily restored factor of $c$ and $G$ to emphasize that this is a quantity
with dimensions of length and can hence be compared with the coupling strength $\ell$.
%
The parameter $T_{\rm EFT}$ was used in both
Refs.~\cite{Nair:2019iur,Perkins:2021mhb} with $\varepsilon = 1/2$,
but here we will assume $\varepsilon \in [0, 1]$.
%
We will say that a bound has been placed on $\ell$, if most of the support of
the posterior distribution function $P(\ell)$ is in the interval $[0, \, T_{\rm EFT}(\varepsilon; \, \mathfrak{m})]$.
%
In practice, this can be quantified by the cumulative distribution function
(CDF) associated to the marginalized posterior distribution $P(\ell)$,
%
\begin{equation}
P(\ell \leqslant x) = \int_{-\infty}^{x} P(x) \, \dd x\,.
\end{equation}
%
We will demand, for instance, that for a bound with 90\% credibility to be meaningfully placed that,
%
\begin{equation}
% \textrm{CDF}(\ell) \leqslant \textrm{CDF}(T_{\rm EFT}(\varepsilon; \, \mathfrak{m})).
P(\ell \leqslant T_{\rm EFT}) \leqslant 0.9 \,.
\end{equation}

%
Another question to consider is: what is the mass $\mathfrak{m}$ that we should use?
%
In Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, which attempted to constrain dCS and shift-symmetric sGB theories with the \emph{inspiral}
signal alone, it was natural to chose the secondary's mass $m_2$ as the most conservative choice, since it is by definition the smallest mass
and hence places the lowest threshold for the validity of either theory as an EFT.
%
In our problem, the answer is not as clear. On the one hand, since we are
interested in the ringdown portion of the signal, it is natural to use the
final mass $M_f$ to compute the threshold $T_{\rm EFT}$.
%
On the other hand, one may argue that the modified gravity theory under study
should be be able to predict a full inspiral, merger, and ringdown of the black
hole binary, and thus the same most conservative $m_2$ should be used.
%
Here we take an agnostic approach and consider both masses, $m_2$ and $M_f$, to
establish $T_{\rm EFT}$, and we will compare how different assumptions yield
different interpretations of the results of our parameter estimations.

There is a second criteria we must also take into account. As already emphasized in Ref.~\cite{Maselli:2019mjd},
the {\sc Parspec} is by construction perturbative. This means that the non-GR deformation parameters are small,
i.e.,
%
\begin{equation}
\gamma \delta \omega^{(j)} \ll 1 \,, \quad \textrm{and} \quad \gamma \delta \tau^{(j)} \ll 1 \,.
\end{equation}
%
for all indices $j$ in the expansion in dimensionless spin $\chi$.
%
We will also construct posterior distributions on the parameters and check if most of their support
is constrained to be much less than unity.

@@

@@

@@

>>>>

\newcommand{\h}{\mathpzc{h}}
\newcommand{\Hhat}{\hat{\mathpzc{H}}}
\newcommand{\B}{\mathpzc{B}}
\newcommand{\hlm}{\mathpzc{h}_{\ell m}}
\newcommand{\xilm}{\xi_{\ell m}}
\newcommand{\Ylm}{{Y}^{-2}_{\ell m}}
\newcommand{\Y}{{Y}^{-2}}
\newcommand{\hc}{h_\times}
\newcommand{\hp}{h_+}
\newcommand{\Fc}{F_\times}
\newcommand{\Fp}{F_+}
\newcommand{\Mf}{M_f}
\newcommand{\cA}{\mathpzc{A}}
\newcommand{\lm}{_{\ell m}}
\newcommand{\deff}{d_\mathrm{eff}}
\newcommand{\rmi}{\mathrm{i}}
\newcommand{\blambda}{\bm{\lambda}}
\newcommand{\btheta}{\bm{\theta}}
\newcommand{\bxi}{\bm{\xi}}
\newcommand{\bxigr}{\bm{\xi}_{\text{GR}}}
\newcommand{\bxingr}{\bm{\xi}_{\text{nGR}}}
\newcommand{\bzeta}{\bm{\zeta}}
\newcommand{\bs}[1]{\bm{\vec{S}_{#1}}}
\newcommand{\Mo}{M_{\odot}}
\newcommand{\FFe}{\mathrm{FF}_\mathrm{eff}}
\newcommand{\FF}{\mathrm{FF}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\rhoopt}{\rho_\mathrm{opt}}
\newcommand{\rhosubopt}{\rho_\mathrm{subopt}}
\newcommand{\fqnm}{f}
\newcommand{\sigmaqnm}{\sigma}
\newcommand{\n}{\mathbf{n}}
\newcommand*{\skymapscale}{0.5}
\newcommand*{\paramestscale}{0.455}
\newcommand{\df}[1]{\delta f_{\text{#1}}}
\newcommand{\dtau}[1]{\delta \tau_{\text{#1}}}
\newcommand{\fngr}[1]{f_{\text{#1}}}
\newcommand{\taungr}[1]{\tau_{\text{#1}}}
\newcommand{\fgr}[1]{f ^{\text{GR}}_{\text{#1}}}
\newcommand{\taugr}[1]{\tau ^{\text{GR}}_{\text{#1}}}
\newcommand{\pSEOB}{\texttt{pSEOBNR}}
\newcommand{\SEOB}{\texttt{SEOBNR}}
\newcommand{\gm}{\mathfrak{m}}

@@

There are two conditions that we must take into account before
we can confidently claim to have placed a constraint, within our model's assumptions,
with our parameter estimation study.

First, as we explained in Sec.~\ref{sec:review_theories}, all theories that we
consider must be considered as an effective field theory, meaning it should be
considered valid only below an energy scale, or equivalently, a length scale.
%
As a cut-off length scale for the validity of the EFT we use,
%
\begin{equation}
\Lambda_{\rm EFT} (\varepsilon; \, \mathfrak{m}) = \varepsilon \,  (G \mathfrak{m} / c^2)\,,
\end{equation}
%
where $\varepsilon$ is a dimensionless number and $\mathfrak{m}$ is the median
value of one of the masses involved in the problem.
%
We also momentarily restored factors of $c$ and $G$ to emphasize that $\Lambda_{\rm EFT}$ has
dimensions of length and hence can be compared to the coupling strength $\ell$.
%
The parameter $\Lambda_{\rm EFT}$ was used
Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr} with $\varepsilon = 1/2$,
but here we  $\varepsilon \in [0, 1]$.
%
We will say that a bound has been placed on $\ell$, if most of the support of
the posterior distribution function $P(\ell)$ is in the interval $[0, \, \Lambda_{\rm EFT}(\varepsilon; \, \mathfrak{m})]$.
%
In practice, this can be quantified by the cumulative distribution function
(CDF) associated to the marginalized posterior distribution $P(\ell)$,
%
\begin{equation}
P(\ell \leqslant \ell') = \int_{-\infty}^{\,\ell'} P(\ell') \, \dd \ell'\,.
\label{eq:def_cdf}
\end{equation}
%
For instance, we will demand that for a bound with 90\% credibility to be meaningfully placed that,
%
\begin{equation}
% \textrm{CDF}(\ell) \leqslant \textrm{CDF}(T_{\rm EFT}(\varepsilon; \, \mathfrak{m})).
P(\ell \leqslant \Lambda_{\rm EFT}) \leqslant 0.9 \,.
\label{eq:eft_bound}
\end{equation}

%
Another question to consider is the following: what is the mass $\mathfrak{m}$ that we should use?
%
In Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}, which attempted to
constrain dCS and shift-symmetric sGB theories with the \emph{inspiral} part of
signal alone, it was natural to chose the secondary's mass $m_2$ as the most
conservative choice, for it is by definition the smallest mass and hence
places the lowest threshold for the validity of either theory as an EFT.
%
In our problem, the answer is not as clear. On the one hand, since we are
interested in the ringdown part of the signal, it is natural to use the
final mass $M_f$ to compute the cut-off $\Lambda_{\rm EFT}$.
%
On the other hand, one may argue that the modified gravity theory under study
should be able to predict a full inspiral, merger, and ringdown of the black
hole binary, and thus the same most conservative $m_2$ should be used.
%
Here we adopt an agnostic view and consider both masses, $m_2$ and $M_f$, to
determine $\Lambda_{\rm EFT}$. We will then compare how different assumptions
yield to different interpretations of the results of our parameter estimations.

There is a second criteria we must also consider.
%
As already emphasized in Ref.~\cite{Maselli:2019mjd}, the {\sc Parspec}
formalism is by construction perturbative. This means that the non-GR
deformation parameters are small, i.e.,
%
\begin{equation}
\gamma \, \delta \omega^{(j)} \ll 1 \,,
\quad \textrm{and} \quad
\gamma \, \delta \tau^{(j)} \ll 1 \,.
\end{equation}
%
for all $j$ of the expansion in dimensionless spin $\chi$.
%
We will also construct posterior distributions on the parameters and check if
most of their support is concentrated to a domain with values much smaller than
unity.

@@

We now turn our attention to dCS gravity.
%
In Fig.~\ref{fig:dCS_exec_sum} we show the marginalized posterior distribution
function of $\ell_{\rm CS}$ for both events GW150914 (top panel) and GW200129
(middle panel), with (dashed line) and without (solid line) spin corrections to
dominant $l = m = 2$,~$n=0$~QNM.
%
Perhaps the most salient feature of these posteriors is that they are not
peaked at $\ell_{\rm CS}$, which may mislead us to believe that we are seeing a
deviation from GR. This is not the case. We must remember that in the {\sc
Parspec} formalism the GR deviations are controlled by $\gamma \, \delta
\omega^{(j)}$~and~$\gamma \, \delta \tau^{(j)}$, and these, indeed, do peak at
zero. (See Fig. xx discussed below.)
%
We also see that in both cases, that the inclusion leading-order-in-spin
correction to the mode, displaces the posteriors towards smaller values of
$\ell_{\rm CS}$. This can be seen more evidently by looking at the location of
posterior peaks.
%
Finally, in the bottom panel, we show the combined result for both events.

In Fig.~\ref{fig:dcs_cdf} we show the cumulative distribution function (CDF) as
evaluates with Eq.~\eqref{eq:def_cdf}, with $\ell' = \Lambda_{\rm EFT}$,
for both GW150914 (top panel) and GW200129 (bottom panel), where $\Lambda_{\rm EFT}$
is calculated with either $\mathfrak{m} = m_2$ (blue curves) or $M_{f}$ (orange curves),
and with $\varepsilon \in [0,\, 1]$.
%
For GW150914, we see that with $\gm = m_2$ that Eq.~\eqref{eq:eft_bound} is not
satisfied unless $\varepsilon \approx 0.9$ (with only $j=0$ corrections) and
$\varepsilon \approx 0.7$ (with both $j=0$ and $1$ corrections).
%
The situation is different if we use $\gm = M_f$. We find that with or without spin corrections
that Eq.~\eqref{eq:eft_bound} can be satisfied with $\varepsilon \leq 1/2$, i.e., below the criteria
used Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
This means that with our model's assumptions and using the remnant's source mass $M_f$ to set the
cut-off scale that we can claim an upper bound
%
\begin{equation}
\ell_{\rm CS} \leqslant \Lambda_{\rm EFT}(1/2; \, M_{f}) = \dots \,.
\quad \textrm{at 90\% credibility},
\end{equation}
%
on dCS gravity and would constitute the first bound on this theory with
gravitational-wave observations alone.
%
Let us consider GW200129 now.

@@

\hs{Try to draw some conclusions that apply to all theories.}

@@

\hs{Say that the fact we spin correction look important reinforces the
importance of calculation QNMs of spinning black holes beyond-GR.}

\hs{Say it is important to go beyond mode calculation and actually understand
how the QNM appear in the GW signal.}
